Gametic Selection, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-01-17 LaTeX/Turnover.tex
added beautiful appendix and updated model section up to characteristic polynomial.

\begin{abstract}
Sex determination systems are remarkably dynamic; many studied taxa display transitions of sex-determining genes between chromosomes or the evolution of new sex-determining systems. 
%Predominant theories in which new sex-determining systems are favoured by selection involve sex ratio selection or fitness differences between sexes (e.g., sexually antagonistic selection). 
Here, we utilize population genetic models to study the spread of novel sex-determining systems in the presence of
%where we also include 
%a period of selection during the haploid stage produced by one sex, e.g., pollen or sperm competition. 
haploid gametic selection, e.g., pollen or sperm competition.
Haploid selected loci experience a form of sex-specific selection (because gametic competition occurs predominantly among haploids produced by males) and can cause sex ratios at birth to become biased (because sex ratios are determined by the fertilization success of X- versus Y-bearing pollen/sperm). 
%We find that the evolution of sex determination systems where mothers determine sex at birth (e.g., environmental sex determination where sex is determined at birth) is influenced by classic Fisherian sex ratio selection. {\color{red}(Maybe not true???)}
%However, notably, we find that the spread of new genetic sex determination systems is not affected by sex ratio biases that are caused by gametic selection because sex ratios become biased after parental provisioning has occurred (even if pollen/sperm competition occurs within the mother). 
Notably, we find that the spread of new genetic sex determination systems is not affected by sex ratio biases that are caused by gametic selection because sex ratios become biased after parental provisioning has occurred (even if pollen/sperm competition occurs within the mother). 
In addition, we find that linkage of an ancestral sex chromosome to a locus under haploid selection can favour transitions between male and female heterogamety (e.g., XY to ZW), which is not the case for any forms of diploid sex specific selection (e.g., sexually antagonistic selection).
During these transitions, new sex-determining alleles spread despite breaking up favourable associations that build up between ancestral sex-determining loci and selected loci, reducing population mean fitness. 
Furthermore, a period of selection among haploids can favour the stable maintenance of polymorphic sex determination systems. 
Thus, our models offer several new insights to be explored as information about sex determination in non-model taxa accumulates.

@@

%genetic system
We consider a three locus model.
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a region under selection, with alleles $A$ and $a$.
And locus \textbf{M} is a novel sex-determining region, with alleles $M$ and $m$.
With genotype $MM$ at locus \textbf{M}, the sex of a zygote is determined by the genotype at locus \textbf{X} ($XX$ become females and $XY$ become males, or $ZW$ become females and $ZZ$ become males).
With at least one $m$ allele at locus \textbf{M}, a zygote develops as a female with probability $k$ and as a male with probability $1-k$.
With $k=0$, the novel sex determiner is a masculinizer (i.e., a neo-Y) and with $k=1$ the novel sex determiner is a feminizer (i.e., a neo-W).
With intermediate $k$, locus \textbf{M} is interpreted as an offspring controlled environmental sex-determining region.
Finally, we also analyze a model of maternally-controlled environmental sex-determination (ESD), where mothers with at least one $m$ allele produce daughters with probability $k$.

%life-cycle
The life-cycle begins with competition between haploid gametes/gametophytes (hereafter gametes) from each sex, where selection depends on the sex of the diploid they came from and their allele at the \textbf{A} locus.
Gametes from males then pair randomly with gametes from females.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (and the \textbf{M} genotype of their mother in the case of maternal control).
Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
This is followed by meiosis with recombination and sex-specific meiotic drive. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$. %is chi really the prob of rec in our recursion equations?
Any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$.
We model meiotic drive at the viability locus only; individuals of sex $i$ who are heterozygous at the \textbf{A} locus produce haploids bearing allele $A$ with probability $\alpha^i$.
We track the frequency of haploid genotypes produced by each sex from one generation to the next (recursion equations in Sup.\ Mat.).

%what were going to do, briefly
With allele $M$ fixed at the \textbf{M} locus, sex is determined by locus \textbf{X} and an equilibrium is reached at locus \textbf{A}. %where, importantly, locus \textbf{A} can be stably polymorphic. %the dynamics can be described by the frequency of $A$ in gametes/gametophytes from the homogametic sex and the frequencies of $A$ in the two types of gametes/gametophytes from the heterogametic sex.
We then examine the ability of a rare, novel sex-determiner, $m$, to increase in frequency from this equilibrium.
Numerical simulations are used to examine when an invading mutation goes to fixation.

@@

\subsection*{Resident equilibrium and stability}
With allele $M$ fixed, we follow the dynamics of the frequency of $A$ in gametes from the homogametic sex (e.g., in eggs from an XX female, $p^f_X$), the frequencies of $A$ in the two types of gametes from the heterogametic sex (e.g., in sperm from an XY male that are X-bearing, $p^m_X$, and Y-bearing, $p^m_Y$), and the frequencies of the sex-determining factors in gametes from the heterogametic sex (e.g., the frequency of Y in sperm, $q$).
Assuming selection and meiotic drive are weak relative to recombination, the differences in the frequencies of $A$ in each type of gamete are small, as is the bias in the sex-determining factor from the heterogametic sex, and we can solve for the mean frequency of $A$ across all types ($p_A$), the difference in the frequencies of $A$ between two of the three types, and the bias in the frequency of the sex-determining factor, to first order in selection.
Linear stability analysis can then be used to determine the stability of this equilibrium.
Without haploid selection or meiotic drive our results reduce to those of \cite{vanDoorn:2007eu}. %when $k=0$ (neo-$Y$ invading a $ZW$ system) or when $k=1$ (neo-$W$ invading an $XY$ system) (\textcolor{red}{CHECK}).
However, with haploid selection or meiotic drive a stable polymorphism at locus \textbf{A} no longer requires sexually antagonistic selection. %, as it can also be achieved by ploidally-antagonistic selection.

\subsection*{Sex chromosome turnover}
The spread of a rare mutant $m$ at the \textbf{M} locus in such a population is determined by the leading eigenvalue, $\lambda$, of the system described by the equations for the next generation frequency of haploid genotypes with the mutation. %, evaluated at the above equilibrium.
Below we present the analytic results for invasion into an $XY$ system assuming no competition among gametes from females, no meiotic drive in females, and linear arrangement \textbf{X}\textbf{A}\textbf{M}. %(i.e., $\chi = r R$). %(SAY SOMETHING HERE ABOUT OTHER ARRANGEMENTS?). %equal dominance coefficients in the two sexes,
Equivalent results for invasions into a $ZW$ system are derived by consistently switching the roles of males and females. %\citep{vanDoorn:2010hu}.

%\subsubsection*{Neo-$Y$}

@@

%write in general form without weak selection
A rare, dominant neo-$Y$ ($k=0$) or neo-$W$ ($k=1$) is always expected to invade an ancestral $XY$ system when the average growth rate of the mutant haplotypes ($Am$ and $am$) is positive, $(g_A + g_a)/2 > 0$. %($g_i = where $\lambda_i$ are the two eigenvalues when $R=r=0$; the \textbf{X} locus does not affect the dynamics).
When the growth rates of the mutant haplotypes without recombination ($R=0$) are negative, $g_i^* < 0 \; \forall \; i \in \{A,a\}$, where $g_i < g_i^*$, the new sex-determining allele does not invade.

@@

\section*{Figures}

@@

>>>>

\begin{abstract}
Sex determination systems are remarkably dynamic; many studied taxa display transitions of sex-determining genes between chromosomes or the evolution of new sex-determining systems. 
%Predominant theories in which new sex-determining systems are favoured by selection involve sex ratio selection or fitness differences between sexes (e.g., sexually antagonistic selection). 
Here, we utilize population genetic models to study the spread of novel sex-determiners in systems with 
%a period of selection during the haploid stage produced by one sex, e.g., pollen or sperm competition. 
haploid gametic selection, e.g., pollen or sperm competition.
Haploid selected loci experience a form of sex-specific selection (because gametic competition occurs predominantly among haploids produced by males) and can cause sex ratios at birth to become biased (because sex ratios are determined by the fertilization success of X- versus Y-bearing pollen/sperm). 
%We find that the evolution of sex determination systems where mothers determine sex at birth (e.g., environmental sex determination where sex is determined at birth) is influenced by classic Fisherian sex ratio selection. {\color{red}(Maybe not true???)}
%However, notably, we find that the spread of new genetic sex determination systems is not affected by sex ratio biases that are caused by gametic selection because sex ratios become biased after parental provisioning has occurred (even if pollen/sperm competition occurs within the mother). 
Notably, we find that the spread of new genetic sex determination systems is not affected by sex ratio biases that are caused by gametic selection because sex ratios become biased after parental provisioning has occurred (even if pollen/sperm competition occurs within the mother). 
In addition, we find that linkage of an ancestral sex chromosome to a locus under haploid selection can favour transitions between male and female heterogamety (e.g., XY to ZW), which is not the case for any forms of diploid sex specific selection (e.g., sexually antagonistic selection).
During these transitions, new sex-determining alleles spread despite breaking up favourable associations that build up between ancestral sex-determining loci and selected loci, reducing population mean fitness. 
Furthermore, a period of selection among haploids can favour the stable maintenance of polymorphic sex determination systems. 
Thus, our models offer several new insights to be explored as information about sex determination in non-model taxa accumulates.

@@

%genetic system
We consider the transition between an ancestral and novel sex determination systems using a three locus model.
Locus \textbf{X} is the ancestral sex-determining region, with alleles $X$ and $Y$ (or $Z$ and $W$).
Locus \textbf{A} is a locus under selection, with alleles $A$ and $a$.
Locus \textbf{M} is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, \textbf{X} ($XX$ become females and $XY$ become males, or $ZW$ become females and $ZZ$ become males). 
To evaluate the evolution of new sex-determination systems, we consider the invasion, fixation, maintenance, and/or loss of novel sex-determining alleles ($m$) at the \textbf{M} locus. 
%We only do invasion analytically though (is this sentence still ok?)
We assume that the \textbf{M} locus is dominant over the \textbf{X} locus such that zygotes with at least one $m$ allele at locus \textbf{M}, a zygote develops as a female with probability $k$ and as a male with probability $1-k$, regardless of the \textbf{X} locus genotype.
With $k=0$, the $m$ allele is a masculinizer (i.e., a neo-Y) and with $k=1$ the $m$ allele is a feminizer (i.e., a neo-W).
With intermediate $k$, the $m$ allele confers environmental sex determination such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
Finally, we also analyze a model of maternally-controlled environmental sex-determination (ESD), where mothers with at least one $m$ allele produce daughters with probability $k$.

%life-cycle
In each generation, we census the genotype frequencies in male and female gametes/gametophytes (hereafter gametes) before haploid competition (recursion equations in Sup.\ Mat.). 
First, competition occurs among male gametes (sperm/pollen competition) and among female gametes (egg/ovule competition). 
Selection during haploid competition depends on the \textbf{A} locus genotype, fitnesses are $w_{A}^m$ and $w_{a}^m$ for male gametes and $w_{A}^f$ and $w_{a}^f$ for female gametes, see table \ref{tab:fitnesstable}. 
Random mating then occurs between male and female gametes.
The resulting zygotes develop as males or females, depending on their genotypes at the \textbf{X} and \textbf{M} loci (and the \textbf{M} genotype of their mother in the case of maternal control) as described above.
Diploid males and females then experience selection, male fitness is given by $w_{h}^{m}$ and female fitness by $w_{h}^{f}$, where $h$ is the genotype at the \textbf{A} locus ($h \in AA, Aa, aa$).  
%Diploids compete with others of the same sex, where selection depends on the sex of the individual and its genotype at the \textbf{A} locus.
The next generation of gametes are then produced by meiosis, during which recombination and sex-specific meiotic drive can occur. 
Recombination occurs between loci \textbf{X} and \textbf{A} with probability $r$, between loci \textbf{A} and \textbf{M} with probability $R$, and between loci \textbf{X} and \textbf{M} with probability $\chi$.
Therefore, any order of the loci can be modelled with appropriate choices of $r$, $R$, and $\chi$ (see Table \ref{tab:chisubstitutions}). 
Males/females that are heterozygous at the \textbf{A} locus experience meiotic drive; $Aa$ heterozgotes of sex $d$ produce gametes bearing allele $A$ with probability $\alpha^d$. 
Thus, the \textbf{A} locus experience sex-specific haploid competition, diploid selection and/or meiotic drive.

\begin{table}[ht]
\centering
\smallskip
\caption{Fitness of different genotypes in sex $d$}
\begin{tabular}{l c }
\hline\hline
  Genotype & Fitness \\ [0.5ex] \hline
  Haploid Competition & \\
  A & $w_{A}^d = 1+t^d$ \\
  a & $w_{a}^d = 1$ \\
  Diploid Selection & \\
  AA & $w_{AA}^d = 1+ s^d$ \\
  Aa & $w_{Aa}^d = 1+h^k s^d$ \\
  aa & $w_{aa}^d = 1$ \\
  \hline \hline
  \label{tab:fitnesstable}
 \end{tabular}
\end{table}

%Info moved above
%what were going to do, briefly
%With allele $M$ fixed at the \textbf{M} locus, sex is determined by locus \textbf{X} and an equilibrium is reached at locus \textbf{A}. %where, importantly, locus \textbf{A} can be stably polymorphic. %the dynamics can be described by the frequency of $A$ in gametes/gametophytes from the homogametic sex and the frequencies of $A$ in the two types of gametes/gametophytes from the heterogametic sex.
%We then examine the ability of a rare, novel sex-determiner, $m$, to increase in frequency from this equilibrium.
%Numerical simulations are used to examine when an invading mutation goes to fixation.

@@

The only asymmetry between males and females in our model is that, under the ancestral sex determination system, males develop with genotype $XY$ (or $ZZ$) and females with genotype $XX$ (or $ZW$). 
Therefore, without loss of generality, we primarily present results for ancestral $XY$ sex determination. 
Ancestral $ZW$ sex determination can be considered by changing the notation such that $X$ becomes $Z$, $Y$ becomes $W$ and the labelling of male and female selection terms are reversed.

%I think this should be moved below because we have some 'results' where we don't assume weak selection but we only have the equilibrium calculated analytically for weak selection. 
\subsection*{Resident equilibrium and stability}
In the resident population (allele $M$ fixed), we follow the frequency of $A$ in female gametes (eggs) from an XX female, $p^f_X$, and in X-bearing, $p^m_X$, and Y-bearing, $p^m_Y$, male gametes (sperm). 
We also track the total frequency of Y-bearing male gametes (sperm), $q$, which may deviate from $1/2$ due to meiotic drive in males.

Various forms of selection can maintain a polymorphism at the \textbf{A} locus, including sexually antagonistic selection, overdominance and conflicts between diploid selection and selection upon haploid genotypes \citep[ploidally antagonistic selection,][]{Immler:2012tl} or a combination of these selective regimes. 
Here, we assume that selection and meiotic drive are weak relative to recombination ($s^k$ and $t^k$ of order $\epsilon$ and $\alpha^k=1/2+\alpha_{\Delta}^k$, where $\alpha_{\Delta}^k$ is of order $\epsilon$). 
The maintenance of a polymorphism at the \textbf{A} locus then requires that

\begin{equation}
\begin{split}
0&<-((1-h^f)s^f +(1-h^m) s^m + t^f +t^m + 2\alpha_{\Delta}^f+2\alpha_{\Delta}^m)\\
%
0&<(h^f s^f +h^m s^m + t^f +t^m + 2\alpha_{\Delta}^f+2\alpha_{\Delta}^m).
\end{split}
\end{equation}

Given that a polymorphism is maintained at the \textbf{A} locus by selection, with weak selection and drive, to leading order, the frequencies of $A$ in each type of gamete are the same ($\hat{p}^f_X=\hat{p}^m_X=\hat{p}^m_Y=\bar{p}$) and given by

\begin{equation}
\bar{p}=\frac{h^f s^f + h^m s^m +t^f+t^m+2\alpha_{\Delta}^f+2\alpha_{\Delta}^m}
{(2h^f-1)s^f+(2h^m-1)s^m}
+O(\epsilon)
.
\label{eq:pAve}
\end{equation}

\noindent
Differences in frequency between gamete types are of order $\epsilon$ to leading order and given by

\begin{equation}
\begin{split}
\hat{p}^m_X-\hat{p}^f_X&=V_{A}\big{(}C^m - C^f \big{)}
+O(\epsilon^2)\\
%
\hat{p}^m_Y-\hat{p}^f_X&=V_{A}\big{(}C^m - C^f+(1-2r)(t^m-t^f)\big{)}/2r
+O(\epsilon^2)\\
%
\hat{p}^m_Y-\hat{p}^m_X&=V_{A}\big{(}C^m - C^f+(t^m-t^f)\big{)}(1-2r)/2r
+O(\epsilon^2)
\end{split}
\end{equation}

\noindent
where $V_{A}=\bar{p}(1-\bar{p})$ is the variance in the frequency of $A$ and $C^d=(\bar{p}+(1-2\bar{p})h^d) s^d +2\alpha_{\Delta}^d$. 
The frequency of Y among male gametes is proportional to the difference in $A$ allele frequency on X- and Y-bearing male gametes, $q=\alpha_{\Delta}^m(\hat{p}^m_Y-\hat{p}^m_X)+O(\epsilon^3)$.
Without haploid competition or drive ($\alpha_{\Delta}^d=t^d=0$), these results reduce to those of \citet{vanDoorn:2007eu}.

% the differences in the frequencies of $A$ in each type of gamete are small, as is the bias in the sex-determining factor from the heterogametic sex, and we can solve for the mean frequency of $A$ across all types ($p_A$), the difference in the frequencies of $A$ between two of the three types, and the bias in the frequency of the sex-determining factor, to first order in selection.
%Linear stability analysis can then be used to determine the stability of this equilibrium.
%Without haploid selection or meiotic drive our results reduce to those of \cite{vanDoorn:2007eu}. %when $k=0$ (neo-$Y$ invading a $ZW$ system) or when $k=1$ (neo-$W$ invading an $XY$ system) (\textcolor{red}{CHECK}).
%However, with haploid selection or meiotic drive a stable polymorphism at locus \textbf{A} no longer requires sexually antagonistic selection. %, as it can also be achieved by ploidally-antagonistic selection.

\subsection*{Sex chromosome turnover}
The evolution of a new sex determination system requires that a rare mutant, $m$, at the novel sex-determining locus  increases in frequency when rare. 
The spread of a rare mutant $m$ at the \textbf{M} locus is determined by the leading eigenvalue, $\lambda$, of the system described by the equations for the next generation frequency of haploid genotypes with the mutation, (\ref{eq:recursions}c), (\ref{eq:recursions}d), (\ref{eq:recursions}g), (\ref{eq:recursions}h). %, evaluated at the above equilibrium.
%Here, we assume loci are arranged in the order \textbf{X}\textbf{A}\textbf{M} (i.e., $\chi=(1-r)R+(1-R)r$). 
%Invasion is numerically evaluated for different locus arrangements in figure \ref{fig:InvasionVsCentiMorgans}, with qualitatively similar results. %(i.e., $\chi = r R$). %(SAY SOMETHING HERE ABOUT OTHER ARRANGEMENTS?). %equal dominance coefficients in the two sexes,

\subsubsection*{Neo-$Y$}

Invasion of a rare, dominant neo-$Y$ ($k=0$) is determined by the largest eigenvalue that solves the characteristic polynomial

\begin{equation}
a \lambda^2+ b \lambda + c =0
\label{eq:charpoly_neoY}
\end{equation}

@@

%<<<<<<< HEAD
%\subsubsection*{Neo-$Y$}
%
%The growth rate of a rare, dominant neo-$Y$ region ($k=0$) is
%\begin{equation}
%\lambda_{Y,XY} = 1 + V_A \frac{(r - R)}{r R} \frac{s_f^2}{\left( s_f + s_m \right)^2} t_m^2,
%\end{equation}
%where $V_A = p_A(1-p_A)$ is the variance at the \textbf{A} locus, $s_f$ and $s_m$ are the respective selection coefficients for $A$ in diploid females and males, and $t_m$ is the selection coefficient for $A$ in gametes/gametophytes from males.
%The neo-Y can therefore invade whenever it is in tighter linkage with the selected locus than is the ancestral sex-determining locus, $r>R$, provided locus \textbf{A} is polymorphic ($V_A>0$) and there is selection among both females ($s_f\neq0$) and gametes/gametophytes from males ($t_m\neq0$).
%=======
%>>>>>>> c89cd9e1d25a76e5bffdd1ee7574884f24c08070
%write in general form without weak selection
A rare, dominant neo-$Y$ ($k=0$) or neo-$W$ ($k=1$) is always expected to invade an ancestral $XY$ system when the average growth rate of the mutant haplotypes ($Am$ and $am$) is positive, $(g_A + g_a)/2 > 0$. %($g_i = where $\lambda_i$ are the two eigenvalues when $R=r=0$; the \textbf{X} locus does not affect the dynamics).
When the growth rates of the mutant haplotypes without recombination ($R=0$) are negative, $g_i^* < 0 \; \forall \; i \in \{A,a\}$, where $g_i < g_i^*$, the new sex-determining allele does not invade.

@@

\newpage
\section*{Figures}

@@

\setcounter{equation}{0}
\renewcommand{\theequation}{S.\arabic{equation}}
\setcounter{figure}{0}
\renewcommand{\thefigure}{S.\arabic{figure}}
\setcounter{table}{0}
\renewcommand{\thetable}{S.\arabic{table}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

\section*{Appendix}

\subsection*{Recursion Equations}
\label{app:recurs}

In each generation we census the genotype frequencies in male and female gametes/gametophytes (hereafter, gametes) before haploid competition. 
Before haploid competition, the frequencies of X-bearing male and female gametes are given by $X_{i}^{m}$ and $X_{i}^{f}$ and the frequencies of Y-bearing gametes are given by $Y_{i}^{m}$ and $Y_{i}^{f}$ where the index $i$ specifies genotypes $MA=1$, $Ma=2$, $mA=3$, and $ma=4$. 
Competition then occurs among gametes of the same sex (e.g., among eggs and among sperm separately) according to the \textbf{A} locus allele, $g$ ($g\in A$, $a$, see Table \ref{tab:fitnesstable}), carried by individuals with genotype $i$.
The genotype frequencies after haploid competition are $X_{i}^{d,s}= w_{g}X_{i}^{d}/\bar{w}_{H}^{d}$ and $Y_{i}^{d,s}= w_{g}Y_{i}^{d}/\bar{w}_{H}^{d}$, where $\bar{w}_{H}^{d}=\sum_{i=1}^{4} w_{g}X_{i}^{d}+w_{g}Y_{i}^{d}$ is the mean fitness of male ($d=m$) or female ($d=f$) gametes. 
Random mating then occurs between gametes to produce diploid zygotes with genotype $ij$ at the \textbf{A} and \textbf{M} loci, such that $XX$ zygotes are denoted $xx_{ij}$, $XY$ zygotes are $xy_{ij}$, and $YY$ zygotes are $yy_{ij}$. 
In $XX$ and $YY$ zygotes, individuals with genotype $ij$ are equivalent to those with genotype $ji$. 
For simplicity, we denote the frequency of genotype $ij$ in $XX$ and $YY$ zygotes to the average of these frequencies, $xx_{ij}=(X_{i}^{f,s}X_{j}^{m,s}+X_{j}^{f,s}X_{i}^{m,s})/2$ and $yy_{ij}=(Y_{i}^{f,s}Y_{j}^{m,s}+Y_{j}^{f,s}Y_{i}^{m,s})/2$.

Denoting the \textbf{M} locus genotype by $b$ ($b\in MM$, $Mm$, $mm$) and the \textbf{X} locus genotype by $c$ ($c\in XX$, $XY$, $YY$), zygotes develop as females with probability $k_{bc}$. 
Therefore, the frequencies of $XX$ females are given by $xx_{ij}^{f}=k_{bc}xx_{ij}$, $XY$ females are given by $xy_{ij}^{f}=k_{bc}xy_{ij}$, and $YY$ females are given by $yy_{ij}^{f}=k_{bc}xy_{ij}$. 
Similarly, $XX$ male frequencies are $xx_{ij}^{m}=(1-k_{bc})xx_{ij}$, $XY$ male frequencies are $xy_{ij}^{m}=(1-k_{bc})xy_{ij}$, and $YY$ males frequencies are $yy_{ij}^{m}=(1-k_{bc})xy_{ij}$.
This notation allows both the ancestral and novel sex-determining regions to determine zygotic sex according to an $XY$ system, a $ZW$ system, or an environmental sex-determining system. 
In addition, we can consider any dominance relationship between the two sex-determining loci. 
Typically, we assume that the ancestral sex-determining system (\textbf{X} locus) is $XY$ ($k_{MMXX}=1$ and $k_{MMXY}=k_{MMYY}=0$) and recessive to a dominant novel sex-determining locus, \textbf{M} ($k_{Mmc}=k_{mmc}=k$).

Selection among diploids then occurs according to the diploid genotype at the \textbf{A} locus, $h$, for an individual of type $ij$ ($h \in AA$, $Aa$, $aa$, see Table \ref{tab:fitnesstable}). 
The diploid frequencies after selection in sex $d$ are given by $xx_{ij}^{d,s}=w_{h}^{d} xx_{ij}/\bar{w}^{d}$, $xy_{ij}^{d,s}=w_{h}^{d} xy_{ij}/\bar{w}^{d}$, and $yy_{ij}^{d,s}=w_{h}^{d} yy_{ij}/\bar{w}^{d}$, where $\bar{w}^{d}= \sum_{i=1}^{4}\sum_{j=1}^{4}w_{h}^{d}xx_{ij}+w_{h}^{d}xy_{ij}+w_{h}^{d}yy_{ij}$ is the mean fitness of individuals of sex $d$.

Finally, these diploids undergo meiosis to produce the next generation of gametes. 
Recombination and sex-specific meiotic drive occur during meiosis.
%We can also assume that recombination is sex specific and/or affected by the M locus - but generally we don't so I just describe $R=R_{f}=R, r_{MM,d}=r_{Mm,d}=r_{mm,d}=r, \chi_{m}=\chi_{f}=\chi$.
Here, we allow the relative locations of the SDR, \textbf{A}, and \textbf{M} loci to be generic by using three parameters to describe the recombination rates between them. $R$ is the recombination rate between the \textbf{A} locus and the \textbf{M} locus, $\chi$ is the recombination rate between the \textbf{M} locus and the \textbf{X} locus, and $r$ is the recombination rate between the \textbf{A} locus and the \textbf{X} locus. 
Table \ref{tab:chisubstitutions} gives substitutions for $\chi$ for defined relative locus locations.
During meiosis in sex $d$, meiotic drive occurs such that, in $Aa$ heterozygotes, a fraction $\alpha_{d}$ of gametes produced carry the $A$ allele and $(1-\alpha^d)$ carry the $a$ allele.

\begin{table}[ht]
\centering
\smallskip
\caption{$\chi$ substitutions for different loci orders (assuming no interference)}
\begin{tabular}{l l}
\hline\hline
  Order of loci &   \\ [0.5ex] \hline
  SDR-A-M & $\chi=R(1-r)+r(1-R)$  \\
  SDR-M-A & $\chi=(r-R)/(1-2R)$ \\
  A-SDR-M & $\chi=(R-r)/(1-2r)$ \\
  \hline \hline
  \label{tab:chisubstitutions}
 \end{tabular}
\end{table}

Among gametes from sex $d$ (sperm/pollen when $d=m$, eggs/ovules when $d=f$), the frequency of haplotypes (before haploid competition) in the next generation are given by

\begin{subequations}
\begin{equation}
\begin{split}
X_{MA}^{d'}=&xx_{11}^{d,s}+xx_{13}^{d,s}/2+(xx_{12}^{d,s}+xx_{14}^{d,s})\alpha^d\\
&-R(xx_{14}^{d,s}-xx_{23}^{d,s}) \alpha^d\\
&+(xy_{11}^{d,s}+xy_{13}^{d,s})/2+(xy_{12}^{d,s}+xy_{14}^{d,s})\alpha^d\\
&- r(xy_{12}^{d,s}-xy_{21}^{d,s})\alpha^d - \chi(xy_{13}^{d,s}-xy_{31}^{d,s})/2\\
&+\big{\{}-(R+r+\chi)xy_{14}^{d,s} +(r+\chi-R)xy_{41}^{d,s}\\
&+(R+r-\chi) xy_{23}^{d,s}+(R+\chi-r)xy_{32}^{d,s}\big{\}}\alpha^d/2
\end{split}
\end{equation}
%
\begin{equation}
\begin{split}
X_{Ma}^{d'}=&xx_{22}^{d,s}+xx_{24}^{d,s}/2+(xx_{12}^{d,s}+xx_{23}^{d,s})\alpha^d\\
&-R(xx_{23}^{d,s}-xx_{14}^{d,s}) \alpha^d\\
&(xy_{22}^{d,s}+xy_{24}^{d,s})/2+(xy_{21}^{d,s}+xy_{23}^{d,s})(1-\alpha^d)\\
&- r(xy_{21}^{d,s}-xy_{12}^{d,s})(1-\alpha^d) - \chi(xy_{24}^{d,s}-xy_{42}^{d,s})/2\\
&+\big{\{}-(R+r+\chi)xy_{23}^{d,s}+(r+\chi-R)xy_{32}^{d,s}\\
&+(R+r-\chi) xy_{14}^{d,s}+(R+\chi-r)xy_{41}^{d,s}\big{\}}(1-\alpha^d)/2
\end{split}
\end{equation}
%
\begin{equation}
\begin{split}
X_{mA}^{d'}=&xx_{33}^{d,s}+xx_{13}^{d,s}/2+(xx_{23}^{d,s}+xx_{34}^{d,s})\alpha^d\\
&-R(xx_{23}^{d,s}-xx_{14}^{d,s}) \alpha^d\\
&(xy_{33}^{d,s}+xy_{31}^{d,s})/2+(xy_{32}^{d,s}+xy_{34}^{d,s})\alpha^d\\
&- r(xy_{34}^{d,s}-xy_{43}^{d,s}) \alpha^d - \chi(xy_{31}^{d,s}-xy_{13}^{d,s})/2\\
&+\big{\{}-(R+r+\chi)xy_{32}^{d,s} +(r+\chi-R)xy_{23}^{d,s}\\
&+(R+r-\chi) xy_{41}^{d,s}+(R+\chi-r)xy_{14}^{d,s}\big{\}} \alpha^d/2
\end{split}
\end{equation}
%
\begin{equation}
\begin{split}
X_{ma}^{d'}=&xx_{44}^{d,s}+xx_{34}^{d,s}/2+(xx_{14}^{d,s}+xx_{24}^{d,s})\alpha^d\\
&-R(xx_{14}^{d,s}-xx_{23}^{d,s}) \alpha^d\\
&(xy_{44}^{d,s}+xy_{42}^{d,s})/2+(xy_{41}^{d,s}+xy_{43}^{d,s})(1-\alpha^d)\\
&- r(xy_{43}^{d,s}-xy_{34}^{d,s})(1-\alpha^d) - \chi(xy_{42}^{d,s}-xy_{24}^{d,s})/2\\
&+\big{\{}-(R+r+\chi)xy_{41}^{d,s}+(r+\chi-R)xy_{14}^{d,s}\\
&+(R+r-\chi) xy_{32}^{d,s}+(R+\chi-r)xy_{23}^{d,s}\big{\}}(1-\alpha^d)/2
\end{split}
\end{equation}
%%%%%%%%%%%%%%%%%%%
%
\begin{equation}
\begin{split}
Y_{MA}^{d'}=&yy_{11}^{d,s}+yy_{13}^{d,s}/2+(yy_{12}^{d,s}+yy_{14}^{d,s})\alpha^d\\
&-R(yy_{14}^{d,s}-yy_{23}^{d,s}) \alpha^d\\
&(xy_{11}^{d,s}+xy_{31}^{d,s})/2+(xy_{21}^{d,s}+xy_{41}^{d,s})\alpha^d\\
&- r(xy_{21}^{d,s}-xy_{12}^{d,s})\alpha^d - \chi(xy_{31}^{d,s}-xy_{13}^{d,s})/2\\
&+\big{\{} - (R+r+\chi)xy_{41}^{d,s} + (r+\chi-R)xy_{14}^{d,s}\\
& + (R+r-\chi)xy_{32}^{d,s} + (R+\chi-r) xy_{23}^{d,s}\big{\}}\alpha^d/2
\end{split}
\end{equation}
%
\begin{equation}
\begin{split}
Y_{Ma}^{d'}=&yy_{22}^{d,s}+yy_{24}^{d,s}/2+(yy_{12}^{d,s}+yy_{23}^{d,s})\alpha^d\\
&-R(yy_{23}^{d,s}-yy_{14}^{d,s}) \alpha^d\\
&(xy_{22}^{d,s}+xy_{42}^{d,s})/2+(xy_{12}^{d,s}+xy_{32}^{d,s})(1-\alpha^d)\\
&- r(xy_{12}^{d,s}-xy_{21}^{d,s})(1-\alpha^d) - \chi(xy_{42}^{d,s}-xy_{24}^{d,s})/2\\
&+\big{\{}-(R+r+\chi)xy_{32}^{d,s} +(r+\chi-R) xy_{23}^{d,s}\\
&+(R+r-\chi)xy_{41}^{d,s}+(R+\chi-r)xy_{14}^{d,s}\big{\}}(1-\alpha^d)/2
\end{split}
\end{equation}
%
\begin{equation}
\begin{split}
Y_{mA}^{d'}=&yy_{33}^{d,s}+yy_{13}^{d,s}/2+(yy_{23}^{d,s}+yy_{34}^{d,s})\alpha^d\\
&-R(yy_{23}^{d,s}-yy_{14}^{d,s}) \alpha^d\\
&(xy_{33}^{d,s}+xy_{13}^{d,s})/2+(xy_{23}^{d,s}+xy_{43}^{d,s})\alpha^d\\
&- r(xy_{43}^{d,s}-xy_{34}^{d,s})\alpha^d - \chi(xy_{13}^{d,s}-xy_{31}^{d,s})/2\\
&+\big{\{}-(R+r+\chi)xy_{23}^{d,s} +(r+\chi-R)xy_{32}^{d,s}\\
&+(R+r-\chi) xy_{14}^{d,s} + (R+\chi-r) xy_{41}^{d,s}\big{\}}\alpha^d/2
\end{split}
\end{equation}
%
\begin{equation}
\begin{split}
Y_{ma}^{d'}=&yy_{44}^{d,s}+yy_{34}^{d,s}/2+(yy_{14}^{d,s}+yy_{24}^{d,s})\alpha^d\\
&-R(yy_{14}^{d,s}-yy_{23}^{d,s}) \alpha^d\\
&(xy_{44}^{d,s}+xy_{24}^{d,s})/2+(xy_{14}^{d,s}+xy_{34}^{d,s})(1-\alpha^d)\\
&- r(xy_{34}^{d,s}-xy_{43}^{d,s})(1-\alpha^d) - \chi(xy_{24}^{d,s}-xy_{42}^{d,s})/2\\
&+\big{\{}-(R+r+\chi) xy_{14}^{d,s} + (r+\chi-R)xy_{41}^{d,s}\\
&+(R+r-\chi) xy_{23}^{d,s} + (R+\chi-r) xy_{32}^{d,s}\big{\}}(1-\alpha^d)/2
\end{split}
\end{equation}
\label{eq:recursions}
\end{subequations}
